---
title: "eda_final_project"
author: "Divya Manoj"
date: "2023-04-17"
output: pdf_document
---
```{r}
install.packages("caret")
```
```{r}
install.packages("randomForest")
```


```{r}
library(readr)
library(caret)
library(randomForest)
```
# Load the Forest Fire dataset into R
```{r}
f = read.csv("C:/Users/manoj/Downloads/forestfires.csv")
```

```{r}
f
```

```{r}
f1 = f
```
```{r}
identical(f, f1)  
```
```{r}
f1$area = log(f1$area+1)
```


# Split the dataset into training and testing sets
```{r}
set.seed(123)
train_index = createDataPartition(f1$area, p = 0.8, list = FALSE)
train = f1[train_index, ]
test = f1[-train_index, ]
```

# Train a random forest model
```{r}
rf_model = randomForest(area ~ ., data = train, ntree = 1000)
```

# Evaluate the model performance on the test set
```{r}
predictions = predict(rf_model, newdata = test)
error = caret::RMSE(predictions, test$area)
cat("RMSE:", error, "\n")
```
# Identify the most important predictors of forest fires
```{r}
importance <- importance(rf_model)
varImpPlot(rf_model)
```
```{r}
print(importance)
```
```{r}
complete.lm <- lm(area ~ X + Y + month + day + FFMC + DMC + DC + ISI + temp + RH + wind + rain, data=f1)
summary(complete.lm)
```


```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
